---
title: ''
author: "Dustin L. Hodge"
date: "Monday, October 19, 2015"
output: html_document
---

# Synopsis
(at most 10 completely sentences summarizing analysis)

# Data Processing

Downloading, unpacking, and loading in CSV data file.
```{r cache=TRUE}
fileName <- "StormData.bz2" # setting once to avoid typos
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
                  fileName)
rawStormData <- read.csv(fileName)
rm(fileName) #don't need this variable anymore
```

We're only concerned with storm effects in the United States proper, not the various islands and territories, so it's a good idea to eliminate them from the dataset under analysis. To do this, we subset the raw data based on the provided FIPS codes in column `STATE_`.

```{r}
stateStormData <- rawStormData[rawStormData$STATE_ <= 56,]
```

It's not immediately obvious why the number `56` is what we're using to subset, because there are 50 states. 51 was my first guess (to include the District of Colmbia), but according to the official [FIPS table](http://www.columbia.edu/~sue/state-fips.html), it's 56. (If you look closely, 5 numbers are omitted entirely; I'm sure there's a story there, but that's someone else's project!)

# Results
Questions:
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?