---
title: "Statistical Inference Project"
author: "Dustin L. Hodge"
date: "December 24, 2015"
output: html_document
---

# Overview

This simulation shows the weird magic of the Central Limit Theorem. We'll take the mean of 40 random samples from an exponential distribution (with `lambda = 0.2` and `mean = 1/lambda`), do that 1000 times, and then take the mean of all those sample means.

# Preparing simulations
```{r simulations, warning=FALSE}
library(ggplot2)

means <- NULL
sims <- 1000; n <- 40; lambda = 0.2

for (i in 1:sims) means <- c(means, mean(rexp(n,lambda)))
```

# Sample Mean vs Theoretical Mean

Now let's see what that looks like.

```{r sample means graph}

tMean <- 1/lambda; sMean <- mean(means)

ggplot(as.data.frame(means), aes(x = means)) +
    geom_histogram(binwidth = .1, fill = "#7e7e7e", color = "#d3d3d3") +
    ggtitle("Histogram of simulation values") +
    geom_vline(xintercept = tMean, color = "red") + 
    geom_point(aes(x = sMean, y = 20, size = 2), color = "blue", show_guide = FALSE) +
    annotate("text", x = sMean + .4, y = 20, label = "Sample mean", color = "blue") +
    annotate("text", x = tMean + .5, y = 55, label = "Theoretical mean", color = "red")
```

How about that! The sample mean and the theoretical mean are almost exactly identical. And though the exponential distribution is, well, exponentially distributed, this collection of sample means maps fairly well to a normal distribution. Well, at least a lot more than the original one does!

# Varience and Distribution

To hammer the distribution difference home, let's discuss variance. First we'll check out the variances.

```{r variances}
# theoretical variance is the same as the theoretical mean for exponential distributions
1/lambda

# and we can use the var command to check out the sample variance
var(means)
```

At first we looked at 1000 averages at 40 samples. Now let's see what the distribution of *40000* samples looks like.

```{r exponential graph}
exampleExp <- sort(rexp(40000, lambda))
qplot(y = 1:length(exampleExp), x = exampleExp, xlab = "value", ylab = "n", 
      main = "Example exponential plot")
```

Not even close to normal. And guess what? The mean is *still 5*.

```{r exponential sample mean}
mean(exampleExp)
```

Looking at the two graphs, it's pretty easy to see why the variance is much larger for the sample. In the sample means plot in the first graph, most of the values are centered around the mean, which is the measure of central tendency. In the above graph, you'd be hard-pressed to even notice a central tendency of the points, because *the vast majority of the data is somewhere else*. Remember that next time you base a decision on a Yelp review.